---
title: "Clean up igotu data"
author: "Kaitlyn"
date: "8/24/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}
library(here)
library(dplyr)
library(rgdal)
library(raster)
library(tidyr)
library(adehabitatLT)
```

### Import shapefile for masking

Import the shapefile of HREC. We'll consider this the huntable area for generating random points and mapping risk. Of course, there are some pastures that need to be masked. 
```{r import hrec shapefile, message = F, results ='hide', warning = F}
# read hrec boundary layer
hrec.boundary.orig <- readOGR(here::here("Data/Spatial data/Raw from Alex"), "HREC_boundary")

# change projection to UTM
hrec.boundary <- spTransform(hrec.boundary.orig, "+proj=utm 
                             +north +zone=10 +ellps=WGS84 +units=m")

###Import shape file with mask of areas that do not allow hunting. 
no.hunt.area.orig <- readOGR(here::here("Data/Spatial data/Raw from Alex"), "hunt_zone")

# change projection to UTM
no.hunt.area <- spTransform(no.hunt.area.orig, "+proj=utm 
                             +north +zone=10 +ellps=WGS84 +units=m")

# create polygon with huntable area only (subtract no hunt from entire boundary)
huntable <- hrec.boundary - no.hunt.area

# make sure that worked
plot(huntable, col = "blue") # it did

#Alex: make this in original proj for other spatial needs:
huntable.wgs <- hrec.boundary.orig - no.hunt.area.orig
huntable.wgs <- spTransform(huntable, "+proj=longlat +ellps=WGS84")
plot(huntable.wgs, col = "red")
```


### Import and clean tracks

Bring in tracks
```{r}
wd <- getwd()
filenames <- list.files(path = paste(wd, "/Data/Hunting/igotu_raw/", sep=""), pattern = NULL, all.files = FALSE, full.names = TRUE, recursive = FALSE, ignore.case = FALSE)
read_csv_filename <- function(filename){
    ret <- read.csv(filename)
    ret$Source <- filename
    ret
}
hunter.csvs <- lapply(filenames, FUN=read_csv_filename)
```

Write function for extracting data from CSVs
```{r}
ExtractIgotuData <- function (X){
  ID <- X[,"Track"]
  unit.time <- X[,"Time"]
  unit.date <- X[,"Date"]
  unit.datetime <- paste(unit.date, unit.time, sep="")
  # translates into POSIXct times
  date_time <- as.POSIXct(strptime(unit.datetime, "%m/%d/%y %H:%M:%S"), tz="America/Los_Angeles")
  # rename lat and long
  lat <-X[,"Latitude"]
  lon <-X[,"Longitude"]
  # spit out as a dataframe
  data.frame(ID=ID, date_time=date_time, lat=lat, lon=lon)
}
```

Download and clean data
```{r}
# generate list of extracted igotu data
extracted.list <- lapply(hunter.csvs, FUN=ExtractIgotuData)

# combine into one
igotu_data_all <- do.call("rbind", extracted.list) 

# add the metadata
metadata <- read.csv(here::here("Data/Hunting/igotu_metadata.csv"))
igotu_data_all <- left_join(igotu_data_all, metadata)
```

Crop to huntable zone
```{r}
# create spatial data frame, transform to UTM, and crop to HREC huntable zone
igotu_spdf <- igotu_data_all # copy
coordinates(igotu_spdf) <- ~lon+lat # create spdf
proj4string(igotu_spdf) <- CRS("+proj=longlat +ellps=WGS84")
igotu_spdf <- spTransform(igotu_spdf, CRS("+proj=utm +north +zone=10 +ellps=WGS84 +units=m"))
igotu_spdf <- igotu_spdf[huntable,]
```

Export data
```{r}
# convert back to dataframe
igotu_data_huntable <- as.data.frame(igotu_spdf)

# write for later
write.csv(igotu_data_huntable, "Data/igotu_data_huntable.csv", row.names=F)
```

Make version with 10 minute fix rate
```{r}
# now make version with 10 min fix rate
hunter.ltraj <- as.ltraj(coordinates(igotu_spdf), igotu_spdf$date_time, id = igotu_spdf$ID, typeII = TRUE)
hunter.ltraj.10min <- redisltraj(hunter.ltraj, u = 600, type = "time")

# get back into data frame, copy metadata in again and format column names
igotu_data_huntable_10min <- ld(hunter.ltraj.10min) %>% 
    rename(date_time = date, lon = x, lat = y, ID = id) %>% 
    left_join(metadata)

# write for later
write.csv(igotu_data_huntable_10min, "Data/igotu_data_huntable_10min.csv", row.names=F)
```