---
title: "Explore the data in hand"
author: "Kaitlyn"
date: "8/21/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

### Investigate metadata for parties and tracks

For further analyses, I think it makes sense to only pick a single track from a given hunting party, to avoid pseudoreplication. I also think it makes sense to divide up the successful and unsuccessful hunting parties (and choose the track from the successful hunter). Here, I explore a bit...

```{r}
library(here)
library(dplyr)
library(rgdal)
library(raster)
library(tidyr)
library(ggplot2)
```

Bring in the metadata that I made manually back in early 2019 (consolidated from field sheets.
```{r}
metadata <- read.csv(here::here("Data/Hunting/igotu_metadata.csv"))
names(metadata)
```

Need to fix the "H_Y" records in the "Hunter_nonhunter" category (originally I had H_Y and H_N)
```{r}
for(i in 1:nrow(metadata)) {
    if(metadata$Hunter_nonhunter[[i]] == "H_Y") {
        metadata$Hunter_nonhunter[[i]] <- "H"
    }
}
```

#### Some simple explorations
```{r}
# how many hunting parties?
length(unique(metadata$Party_ID)) # 137

# how many individuals successfully harvested?
count(metadata, Harvest) # 36

# how many parties had more than one harvest? (answer = 7)
(multipleharvests <- metadata %>% 
    filter(Harvest == "Y") %>% 
    count(Party_ID) %>% 
    filter(n > 1) %>% 
    mutate(Party_ID = droplevels(Party_ID)))
```

There are 7 parties in which two individuals harvested animals. 

#### Unsuccessful recovery

For early files, we know whether or not the data were successfully recovered. Looks like 6 tags didn't work (plus there are 70 that we don't have any more information about).
```{r}
count(metadata, Data_recovered) 

filter(metadata, Data_recovered == "N") 

# remove the ones that didn't work
metadata_work <- filter(metadata, Data_recovered != "N")
```

Unfortunately, 080815_17 from party 080815_G harvested a deer but had a failed tag. Could potentially use the other tag from that party.


#### Explore non-hunters
```{r}
# how many nonhunters?
count(metadata, Hunter_nonhunter)

# start by removing those
metadata_hunter <- filter(metadata, Hunter_nonhunter == "H")

# were there any parties for which that was the ONLY tag that we successfully collected? (no)
length(unique(metadata_hunter$Party_ID)) == length(unique(metadata$Party_ID)) 
# true; all good
```

#### Summarize what's going on in each party
```{r}
summary_hunters <- metadata %>% 
  count(Party_ID, Hunter_nonhunter) %>% 
  pivot_wider(id_cols = Party_ID,
              names_from = Hunter_nonhunter,
              values_from = n)

summary_harvest <- metadata %>% 
  count(Party_ID, Harvest) %>% 
  pivot_wider(id_cols = Party_ID,
              names_from = Harvest,
              values_from = n) 

summary <- full_join(summary_hunters, summary_harvest) %>% 
  replace_na(replace = list(NH = 0, N = 0, Y = 0)) %>% 
  rename(No_Harvest = N, Harvest = Y, Hunters = H, NonHunters = NH)
```


### Bring in iGotU tracks

Cleaned on 01-clean-data.Rmd file

```{r}
igotu_data_huntable <- read.csv(here::here("Data/igotu_data_huntable.csv"))
```




### Deep dive into each party

Plot the individual tracks for each of the parties that had multiple kills
```{r}
igotu_data_huntable %>% 
    filter(Party_ID %in% levels(multipleharvests$Party_ID)) %>% 
    ggplot(aes(x = lon, y = lat, col = ID)) +
    geom_point() +
    facet_wrap(~Party_ID)
```


```{r}
party_summary <- function(party) {
  
  # metadata for all igotus in this party
  print(metadata %>% 
    filter(Party_ID == party))
  
  # summary
  print(summary %>% 
    filter(Party_ID == party))

  # plot of all igotus  
  igotu_data_huntable %>% 
    filter(Party_ID == party) %>% 
    ggplot(aes(x = lon, y = lat, col = ID)) +
    geom_point()
}
```

```{r}
lapply(levels(metadata$Party_ID), party_summary)
```

