---
title: "Explore the data in hand"
author: "Kaitlyn"
date: "8/21/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

### Investigate metadata for parties and tracks

For further analyses, I think it makes sense to only pick a single track from a given hunting party, to avoid pseudoreplication. I also think it makes sense to divide up the successful and unsuccessful hunting parties (and choose the track from the successful hunter). Here, I explore a bit...

```{r}
library(here)
library(dplyr)
library(rgdal)
library(raster)
library(tidyr)
```

Bring in the metadata that I made.
```{r}
metadata <- read.csv(here::here("Data/Hunting/igotu_metadata.csv"))
names(metadata)

# how many hunting parties?
length(unique(metadata$Party_ID)) # 137

# how many individuals successfully harvested?
count(metadata, Harvest) # 36

# how many parties had more than one harvest? (answer = 7)
(multipleharvests <- metadata %>% 
    filter(Harvest == "Y") %>% 
    count(Party_ID) %>% 
    filter(n > 1) %>% 
    mutate(Party_ID = droplevels(Party_ID)))
```

There are 7 parties in which two individuals harvested animals. 

For early files, we know whether or not the data were successfully recovered. Looks like 6 tags didn't work (plus there are 70 that we don't have any more information about)
```{r}
count(metadata, Data_recovered) 

# remove the ones that didn't work
metadata_work <- filter(metadata, Data_recovered != "N")
```

Try removing the non-hunters from this list.
```{r}
# need to fix the "H_Y" records in the "Hunter_nonhunter" category (originally I had H_Y and H_N)
for(i in 1:nrow(metadata)) {
    if(metadata$Hunter_nonhunter[[i]] == "H_Y") {
        metadata$Hunter_nonhunter[[i]] <- "H"
    }
}

# how many nonhunters?
count(metadata, Hunter_nonhunter)

# start by removing those
metadata_hunter <- filter(metadata, Hunter_nonhunter == "H")

# were there any parties for which that was the ONLY tag that we successfully collected? (no)
length(unique(metadata_hunter$Party_ID)) == length(unique(metadata$Party_ID)) 
# true; all good
```

Summarize what's going on in each party
```{r}
summary_hunters <- metadata %>% 
  count(Party_ID, Hunter_nonhunter) %>% 
  pivot_wider(id_cols = Party_ID,
              names_from = Hunter_nonhunter,
              values_from = n)

summary_harvest <- metadata %>% 
  count(Party_ID, Harvest) %>% 
  pivot_wider(id_cols = Party_ID,
              names_from = Harvest,
              values_from = n) 

summary <- full_join(summary_hunters, summary_harvest) %>% 
  replace_na(replace = list(NH = 0, N = 0, Y = 0)) %>% 
  rename(No_Harvest = N, Harvest = Y, Hunters = H, NonHunters = NH)
```


### Import shapefile for masking

Import the shapefile of HREC. We'll consider this the huntable area for generating random points and mapping risk. Of course, there are some pastures that need to be masked. 
```{r import hrec shapefile, message = F, results ='hide', warning = F}
# read hrec boundary layer
hrec.boundary.orig <- readOGR(here::here("Data/Spatial data/Raw from Alex"), "HREC_boundary")

# change projection to UTM
hrec.boundary <- spTransform(hrec.boundary.orig, "+proj=utm 
                             +north +zone=10 +ellps=WGS84 +units=m")

###Import shape file with mask of areas that do not allow hunting. 
no.hunt.area.orig <- readOGR(here::here("Data/Spatial data/Raw from Alex"), "hunt_zone")

# change projection to UTM
no.hunt.area <- spTransform(no.hunt.area.orig, "+proj=utm 
                             +north +zone=10 +ellps=WGS84 +units=m")

# create polygon with huntable area only (subtract no hunt from entire boundary)
huntable <- hrec.boundary - no.hunt.area

# make sure that worked
plot(huntable, col = "blue") # it did

#Alex: make this in original proj for other spatial needs:
huntable.wgs <- hrec.boundary.orig - no.hunt.area.orig
huntable.wgs <- spTransform(huntable, "+proj=longlat +ellps=WGS84")
plot(huntable.wgs, col = "red")
```



### Deep dive into each party

Plot the individual tracks for each of the parties that had multiple kills
```{r}
igotu_data_huntable %>% 
    filter(Party_ID %in% levels(multipleharvests$Party_ID)) %>% 
    ggplot(aes(x = lon, y = lat, col = ID)) +
    geom_point() +
    facet_wrap(~Party_ID)
```


```{r}
party_summary <- function(party) {
  
  print(party)
  
  # metadata for all igotus in this party
  metadata %>% 
    filter(Party_ID == party)
  
  # summary
  summary %>% 
    filter(Party_ID == party)
  
  # plot of each igotu
  igotu_data_huntable %>% 
    filter(Party_ID == party) %>% 
    ggplot(aes(x = lon, y = lat, col = ID)) +
    geom_point() +
    facet_wrap(~ID)
  
  # plot of all igotus  
  igotu_data_huntable %>% 
    filter(Party_ID == party) %>% 
    ggplot(aes(x = lon, y = lat, col = ID)) +
    geom_point()
}
```

```{r}
lapply(levels(metadata$Party_ID), party_summary)
```

