---
title: "State Space Models (HMM)"
author: "Kaitlyn"
date: "8/21/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

## Explore Hidden Markov Models for Hopland hunter data

```{r}
library(moveHMM)
library(ggplot2)
library(dplyr)
library(sf)
library(tidyr)
```


Load cleaned data.
```{r}
igotu_data <- read.csv("Data/igotu_data_3min_covariates.csv")

# plot one just to see what's up
igotu_data %>% 
  filter(ID == "080815_01") %>% 
  ggplot(aes(x = Longitude, y = Latitude)) +
  geom_point() 
```

Prep data for modeling

```{r}
igotu_data_fewer <- igotu_data %>% 
  dplyr::select(ID, Longitude, Latitude, DateTime,
                rugged49.clean, rugged25.clean, rugged9.clean,
                hq_dist, vegetation.coarser.clean2, view,
                veg.edges.dist.clean,
                grass_120m, chap_120m, wood_120m)

data_hmm <- moveHMM::prepData(igotu_data_fewer, 
                              type="LL", 
                              coordNames=c("Longitude","Latitude"))
head(data_hmm)
```

Remove outliers based on step length (this will remove GPS error & gaps in data collection)

1 mile/hr = 0.0268224 km/min = 0.0804672 km step length (km/3 min)
10 mile/hr = 0.268224 km/min, or 0.804672 km step length (km/3 min)
15 mile/hr = 0.402336 km/min, or 1.207008 km step length (km/3 min)
20 mile/hr = 0.536448 km/min, or 1.609344 km step length (km/3 min)
25 mile/hr = 0.67056 km/min, or 2.01168 km step length (km/3 min)
30 mile/hr = 0.804672 km/min, or 2.414016 km step length (km/3 min)
```{r}
# remove outliers (step length in top 95th percentile) - indicative of GPS error or gaps in data collection
# step length is in km
data_hmm <- data_hmm %>% 
  drop_na(step) # remove 300 step lenghts of 'NA'
hist(data_hmm$step)

# 10mph cutoff - 2% of points (1,619 points)
nrow(filter(data_hmm, step > 0.804672))/nrow(data_hmm)

# 15mph cutoff - 0.4% of points (295 points)
nrow(filter(data_hmm, step > 1.207008))/nrow(data_hmm)

# 20mph cutoff - 0.17% of points (124 points)
nrow(filter(data_hmm, step > 1.609344))/nrow(data_hmm)

# 25mph cutoff - 0.10% of points (76 points)
nrow(filter(data_hmm, step > 2.01168))/nrow(data_hmm)

# 30mph cutoff - 0.06% of points (47 points)
nrow(filter(data_hmm, step > 2.414016))/nrow(data_hmm)

# histogram for only steps >= 10mph
fast_steps <- data_hmm %>% 
  filter(step > 0.804672) %>% 
  mutate(speed = step/3 * 37.2823)
hist(fast_steps$speed)

plot(data_hmm[data_hmm$ID == "080915_02",])

summary(data_hmm)

# LET'S USE 15 mph cut-off since it seems like there is a big drop there
data_hmm <- data_hmm %>% 
  filter(step < 1.207008)
```

Have a look at the zero step lengths
```{r}
# determine proportion of step lengths equal to 0
whichzero <- which(data_hmm$step == 0)
length(whichzero)/nrow(data_hmm)

# 0.19 for 3 min

zeros <- data_hmm %>% 
  filter(step == 0)

ggplot(zeros, aes(x = x, y = y)) + geom_point()

#  data_hmm_copy <- data_hmm
#  data_hmm_copy[is.na(data_hmm_copy)] <- 0
#  
#  # determine which ones follow a zero
#  data_hmm_copy$after_zero <- "NOT_AFTER_ZERO" 
#  for(i in 2:nrow(data_hmm_copy)) {
#    if(data_hmm_copy$step[i] == 0) {
#      data_hmm_copy$after_zero[i] <- "ZERO"
#    } else if(data_hmm_copy$step[i-1] == 0) {
#      data_hmm_copy$after_zero[i] <- "AFTER_ZERO"
#    } else {
#      data_hmm_copy$after_zero[i] <- "NOT_AFTER_ZERO"
#    }
#  }
#  
#  ggplot(data_hmm_copy, aes(x = step)) +
#    geom_histogram() +
#    facet_wrap(~after_zero)
```


Prep & run model
```{r}
hist(data_hmm$step)
hist(data_hmm$angle)

mu0 <- c(0.01, 0.5) # step mean (two parameters: one for each state)
sigma0 <- c(0.01, 0.5) # step SD
zeromass0 <- c(0.50, 0.10) # step zero-mass
stepPar0 <- c(mu0,sigma0,zeromass0)

angleMean0 <- c(pi,0) # angle mean
kappa0 <- c(1,1) # angle concentration
anglePar0 <- c(angleMean0,kappa0)

m <- fitHMM(data=data_hmm, 
            nbStates=2, 
            stepPar0=stepPar0, 
            anglePar0=anglePar0,
            formula = ~view)

m

plot(m)

# proportion of time spent in each state
states <- viterbi(m)
prop.table(table(states)) 

# plot states for a given individual
plotStates(m, animals="080815_01")

plotStates(m)
```

Now try separating successful vs unsuccessful hunters
```{r}
successful <- igotu_data %>% 
  filter(Harvest == "Y") %>% 
  dplyr::select(ID, lon, lat, rugged49.clean, slope, hq_dist, vegetation.coarser.clean2, view)
unsuccessful <- igotu_data %>% 
  filter(Harvest == "N") %>% 
  dplyr::select(ID, lon, lat, rugged49.clean, slope, hq_dist, vegetation.coarser.clean2, view)

data.successful <- moveHMM::prepData(successful, type="LL", coordNames=c("lon","lat"))
data.unsuccessful <- moveHMM::prepData(unsuccessful, type="LL", coordNames=c("lon","lat"))

hist(data.successful$step, col= 'red', breaks = 50, freq = F); hist(data.unsuccessful$step, col = 'blue', add=T, breaks = 50, freq = F)

hist(data.successful$angle, col= 'red', breaks = 50, freq = F); hist(data.unsuccessful$angle, col = 'blue', add=T, breaks = 50, freq = F)
```

#### Successful vs unsuccessful hunters model

With no transition probabilities
```{r}
m.successful <- fitHMM(data=data.successful, 
            nbStates=2, 
            stepPar0=stepPar0, 
            anglePar0=anglePar0)

m.successful
plot(m.successful)

m.unsuccessful <- fitHMM(data=data.unsuccessful, 
            nbStates=2, 
            stepPar0=stepPar0, 
            anglePar0=anglePar0)

m.unsuccessful
plot(m.unsuccessful)

```



#### Three state model
```{r}
# initial parameters
mu0 <- c(0.01, 0.5, 1.0)
sigma0 <- c(0.01, 0.5, 1.0)
zeromass0 <- c(0.40, 0.40, 0.10)
stepPar3 <- c(mu0,sigma0,zeromass0)

angleMean0 <- c(pi,pi,0)
kappa0 <- c(1,1,1)
anglePar3 <- c(angleMean0,kappa0)

m3 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar3, anglePar0=anglePar3)

states <- viterbi(m3)
states

# proportion of time spent in each state
prop.table(table(states)) 

plot(m3)
  
AIC(m, m3)
```

## BPCA

abandoned this after some brief exploration

```{r}
igotu_data

library(bcpa)

X <- as.numeric(coords@coords[1:100,1])
Y <- as.numeric(coords@coords[1:100,2])
Time <- 1:100
mytrack <- MakeTrack(X,Y,Time)
plot(mytrack)

test_igotu <- igotu_data %>% 
  filter(ID == "081416_04")

mytrack <- MakeTrack(test_igotu$lon, test_igotu$lat, 
                     Time = test_igotu$date_time)

plot(mytrack)

igotu.VT <- GetVT(mytrack)
head(igotu.VT)
```

