---
title: "04-hmm-3state"
author: "Kaitlyn"
date: "1/21/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

## Load and clean data

```{r}
library(dplyr)
library(moveHMM)
library(tidyr)
library(ggplot2)
```

Load data & remove outliers.
```{r}
# Load cleaned data
igotu_data <- read.csv("Data/igotu_data_3min_covariates.csv")


# Select columns of interest
igotu_data_fewer <- igotu_data %>% 
    dplyr::select(ID, Party_ID, Longitude, Latitude, DateTime,
                  rugged49.clean, rugged25.clean, rugged9.clean,
                  hq_dist, vegetation.coarser.clean2, view_for_kg_proj,
                  road.dist.clean,
                  grass_120m, chap_120m, wood_120m,
                  Elapsed_Time, Elapsed_Time_Sunrise, Elapsed_Time_SunriseOrSet,
                  Harvest)

# prep data for HMM
data_hmm <- moveHMM::prepData(igotu_data_fewer, 
                              type="LL", 
                              coordNames=c("Longitude","Latitude"))

# remove 300 step lengths of 'NA'
data_hmm <- data_hmm %>% 
    drop_na(step) 

# filter out all steps > 15mph
data_hmm <- data_hmm %>% 
    filter(step < 1.207008)
```

Scale covariates.
```{r}
head(data_hmm)
data_hmm$rugged49_scale <- scale(data_hmm$rugged49.clean)
data_hmm$rugged25_scale <- scale(data_hmm$rugged25.clean)
data_hmm$rugged9_scale <- scale(data_hmm$rugged9.clean)
data_hmm$hq_scale <- scale(data_hmm$hq_dist)
data_hmm$view_scale <- scale(data_hmm$view_for_kg_proj)
data_hmm$road_scale <- scale(data_hmm$road.dist.clean)
data_hmm$grass_scale <- scale(data_hmm$grass_120m)
data_hmm$chap_scale <- scale(data_hmm$chap_120m)
data_hmm$wood_scale <- scale(data_hmm$wood_120m)
data_hmm$time_scale <- scale(data_hmm$Elapsed_Time)
data_hmm$sunrise_scale <- scale(data_hmm$Elapsed_Time_Sunrise)
data_hmm$sunriseset_scale <- scale(data_hmm$Elapsed_Time_SunriseOrSet)
head(data_hmm)
```

Explore correlation between spatial covariates.
```{r}
covariates <- dplyr::select(data_hmm, rugged49_scale, rugged25_scale, rugged9_scale,
                     hq_scale, view_scale, road_scale, grass_scale,
                     chap_scale, wood_scale, time_scale, sunrise_scale, sunriseset_scale)
round(cor(covariates), digits = 2)
```

There is high correlation between:
0.83 = rugged25_scale and rugged49_scale
0.72 = rugged25_scale and rugged9_scale
-0.69 = grass_scale and wood_scale


# THREE STATE MODELS

Determined the initial values from the exploration in 04b-hmm-3state-parameter-select.Rmd
 
```{r}
mu0_3state <- c(0.01, 0.07, 0.41)
sigma0_3state <- c(0.010, 0.045, 0.204)
zeromass0_3state <- c(0.469, 0.047, 0.0028) 
stepPar0_3state <- c(mu0_3state,sigma0_3state,zeromass0_3state)
angleMean0_3state <- c(pi,0,0)
kappa0_3state <- c(0.61,1.20,1.14)
anglePar0_3state <- c(angleMean0_3state,kappa0_3state)
```

### Null model
```{r}
model0 <- fitHMM(data=data_hmm, nbStates=3, 
             stepPar0=stepPar0_3state, anglePar0=anglePar0_3state)
 
states <- viterbi(model0)
states
 
# proportion of time spent in each state
prop.table(table(states)) 
 
plot(model0)
AIC(model0) # 69566.89
```

## Univariate models
 
### Ruggedness (and viewshed)
```{r}
model1 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar0_3state, anglePar0=anglePar0_3state,
                 formula = ~rugged25_scale)
model1 # model summary
AIC(model1) # AIC = 69404.87

model2 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar0_3state, anglePar0=anglePar0_3state,
                 formula = ~rugged49_scale)
model2 # model summary
AIC(model2) # AIC = 69466.68

model3 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar0_3state, anglePar0=anglePar0_3state,
                 formula = ~rugged9_scale)
model3 # model summary
AIC(model3) # AIC = 69331.39 - BEST ruggedness layer
```

### Viewshed
```{r}
model4 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar0_3state, anglePar0=anglePar0_3state,
                 formula = ~view_scale)
model4 # model summary
AIC(model4) # AIC = 69529.46
```

### Distance to road
```{r}
model5 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar0_3state, anglePar0=anglePar0_3state,
                 formula = ~road_scale)
model5 # model summary
AIC(model5) # AIC = 67167.63
```

### Woodland within 120 meters
```{r}
model6 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar0_3state, anglePar0=anglePar0_3state,
                 formula = ~wood_120m)
model6 # model summary
AIC(model6) # AIC = 69476.21
```

### Chaparral within 120 meters
```{r}
model7 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar0_3state, anglePar0=anglePar0_3state,
                 formula = ~chap_120m)
model7 # model summary
AIC(model7) # AIC = 69535.18
```

### Grassland within 120 meters
```{r}
model8 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar0_3state, anglePar0=anglePar0_3state,
                 formula = ~grass_120m)
model8 # model summary
AIC(model8) # AIC = 69530.39
```

### Distance to headquarters
```{r}
model9 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar0_3state, anglePar0=anglePar0_3state,
                 formula = ~hq_dist)
model9 # model summary
AIC(model9) # AIC = 69514.61
```

### Time variables
```{r}
model10 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar0_3state, anglePar0=anglePar0_3state,
                 formula = ~time_scale)
model10 # model summary
AIC(model10) # AIC = 69367.41

model11 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar0_3state, anglePar0=anglePar0_3state,
                 formula = ~sunrise_scale)
model11 # model summary
AIC(model11) # AIC = 69361.25

model12 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar0_3state, anglePar0=anglePar0_3state,
                 formula = ~sunriseset_scale)
model12 # model summary
AIC(model12) # AIC = 69535.25
```


## Two-variable models

NOTE: AIC values for the models below were all calculated (accidentally) using the UNSCALED version of distance to road. But all model exploration suggests that the full model is the best, so going to skip the stepwise selection process and just say that we are going with the full model (aside from highly correlated covariates) since it best represents our hypotheses.

### Distance to road + Ruggedness
```{r}
model2.1 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar0_3state, anglePar0=anglePar0_3state,
                 formula = ~road_scale + rugged9_scale)
model2.1 # model summary
AIC(model2.1) # AIC = 67150.65
```

### Distance to road + Chaparral
```{r}
model2.2 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar0_3state, anglePar0=anglePar0_3state,
                 formula = ~road_scale + chap_scale)
model2.2 # model summary
AIC(model2.2) # AIC = 67168.28
```

### Distance to road + Grassland
```{r}
model2.3 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar0_3state, anglePar0=anglePar0_3state,
                 formula = ~road_scale + grass_scale)
model2.3 # model summary
AIC(model2.3) # AIC = 67168.86
```

### Distance to road + Woodland
```{r}
model2.4 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar0_3state, anglePar0=anglePar0_3state,
                 formula = ~road_scale + wood_scale)
model2.4 # model summary
AIC(model2.4) # AIC = 67166.28
```

### Distance to road + Viewshed
```{r}
model2.5 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar0_3state, anglePar0=anglePar0_3state,
                 formula = ~road_scale + view_scale)
model2.5 # model summary
AIC(model2.5) # AIC = 67138.82
```

### Distance to road + Distance to HQ
```{r}
model2.6 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar0_3state, anglePar0=anglePar0_3state,
                 formula = ~road_scale + hq_scale)
model2.6 # model summary
AIC(model2.6) # AIC = 67133.55
```

### Distance to road + Time - BEST
```{r}
model2.7 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar0_3state, anglePar0=anglePar0_3state,
                 formula = ~road_scale + sunrise_scale)
model2.7 # model summary
AIC(model2.7) # AIC = 67001.23
```


## Three-variable models

### Distance to road + Time + Ruggedness
```{r}
model3.1 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar0_3state, anglePar0=anglePar0_3state,
                 formula = ~road_scale + sunrise_scale + rugged9_scale)
model3.1 # model summary
AIC(model3.1) # AIC = 66986.97
```

### Distance to road + Time + Viewshed
```{r}
model3.2 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar0_3state, anglePar0=anglePar0_3state,
                 formula = ~road_scale + sunrise_scale + view_scale)
model3.2 # model summary
AIC(model3.2) # AIC = 66971.34
```

### Distance to road + Time + Chaparral
```{r}
model3.3 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar0_3state, anglePar0=anglePar0_3state,
                 formula = ~road_scale + sunrise_scale + chap_scale)
model3.3 # model summary
AIC(model3.3) # AIC = 67002.39
```

### Distance to road + Time + Woodland
```{r}
model3.4 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar0_3state, anglePar0=anglePar0_3state,
                 formula = ~road_scale + sunrise_scale + wood_scale)
model3.4 # model summary
AIC(model3.4) # AIC = 67000.24
```

### Distance to road + Time + Grassland
```{r}
model3.5 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar0_3state, anglePar0=anglePar0_3state,
                 formula = ~road_scale + sunrise_scale + grass_scale)
model3.5 # model summary
AIC(model3.5) # AIC = 67001.83
```

### Distance to road + Time + Distance to HQ
```{r}
model3.6 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar0_3state, anglePar0=anglePar0_3state,
                 formula = ~road_scale + sunrise_scale + hq_scale)
model3.6 # model summary
AIC(model3.6) # AIC = 66963.17
```


## Four-variable models

### Distance to road + Time + Distance to HQ + Ruggedness
```{r}
model4.1 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar0_3state, anglePar0=anglePar0_3state,
                 formula = ~road_scale + hq_scale + view_scale + rugged9_scale)
model4.1 # model summary
AIC(model4.1) # AIC = 67075.9
```

### Distance to road + Time + Distance to HQ + Chaparral
```{r}
model4.2 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar0_3state, anglePar0=anglePar0_3state,
                 formula = ~road_scale + hq_scale + view_scale + chap_scale)
model4.2 # model summary
AIC(model4.2) # AIC = 
```

### Distance to road + Time + Distance to HQ + Woodland
```{r}
model4.3 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar0_3state, anglePar0=anglePar0_3state,
                 formula = ~road_scale + hq_scale + view_scale + wood_scale)
model4.3 # model summary
AIC(model4.3) # AIC = 67103.85
```

### Distance to road + Time + Distance to HQ + Grassland
```{r}
model4.4 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar0_3state, anglePar0=anglePar0_3state,
                 formula = ~road_scale + hq_scale + view_scale + grass_scale)
model4.4 # model summary
AIC(model4.4) # AIC = 67099.1
```

### Distance to road + Time + Distance to HQ + Viewshed
```{r}
model4.5 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar0_3state, anglePar0=anglePar0_3state,
                 formula = ~road_scale + hq_scale + view_scale + view_scale)
model4.5 # model summary
AIC(model4.5) # AIC = 67101.3
```


## Five-variable models

### Distance to road + Time + Distance to HQ + Chaparral + Ruggedness
```{r}
model5.1 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar0_3state, anglePar0=anglePar0_3state,
                 formula = ~road_scale + hq_scale + view_scale + chap_scale + rugged9_scale)
model5.1 # model summary
AIC(model5.1) # AIC = 
```

### Distance to road + Time + Distance to HQ + Chaparral + Woodland
```{r}
model5.2 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar0_3state, anglePar0=anglePar0_3state,
                 formula = ~road_scale + hq_scale + view_scale + chap_scale + wood_scale)
model5.2 # model summary
AIC(model5.2) # AIC = 
```

### Distance to road + Time + Distance to HQ + Chaparral + Grassland
```{r}
model5.3 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar0_3state, anglePar0=anglePar0_3state,
                 formula = ~road_scale + hq_scale + view_scale + chap_scale + grass_scale) 
model5.3 # model summary
AIC(model5.3) # AIC = 
```

### Distance to road + Time + Distance to HQ + Chaparral + Viewshed
```{r}
model5.4 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar0_3state, anglePar0=anglePar0_3state,
                 formula = ~road_scale + hq_scale + view_scale + chap_scale + view_scale)
model5.4 # model summary
AIC(model5.4) # AIC = 
```


## Six-variable models

### Distance to road + Distance to HQ + Viewshed + Woodland + Ruggedness + Chaparral
```{r}
model6.1 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar0_3state, anglePar0=anglePar0_3state,
                 formula = ~)
model6.1 # model summary
AIC(model6.1) # AIC = 
```

### Distance to road + Distance to HQ + Viewshed + Woodland + Ruggedness + Time
```{r}
model6.2 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar0_3state, anglePar0=anglePar0_3state,
                 formula = ~)
model6.2 # model summary
AIC(model6.2) # AIC = 
```

## Seven-variable models

### Distance to road + Distance to HQ + Viewshed + Woodland + Ruggedness + Chaparral + Time
```{r}
model7.1 <- fitHMM(data=data_hmm, nbStates=3, stepPar0=stepPar0_3state, anglePar0=anglePar0_3state,
                 formula = ~road_scale + hq_scale + view_scale + wood_scale + rugged9_scale + chap_scale + sunrise_scale)
model7.1 # model summary
AIC(model7.1) # AIC = 66883.21
```

