---
title: "State Space Models (HMM)"
author: "Kaitlyn"
date: "8/21/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

## Load and clean data

```{r}
library(dplyr)
library(moveHMM)
library(tidyr)
library(ggplot2)
```

Load data & remove outliers.
```{r}
# Load cleaned data
igotu_data <- read.csv("Data/igotu_data_3min_covariates.csv")

# Select columns of interest
igotu_data_fewer <- igotu_data %>% 
    dplyr::select(ID, Party_ID, Longitude, Latitude, DateTime,
                  rugged49.clean, rugged25.clean, rugged9.clean,
                  hq_dist, vegetation.coarser.clean2, view_for_kg_proj,
                  road.dist.clean,
                  grass_120m, chap_120m, wood_120m)

# prep data for HMM
data_hmm <- moveHMM::prepData(igotu_data_fewer, 
                              type="LL", 
                              coordNames=c("Longitude","Latitude"))

# remove 300 step lengths of 'NA'
data_hmm <- data_hmm %>% 
    drop_na(step) 

# filter out all steps > 15mph
data_hmm <- data_hmm %>% 
    filter(step < 1.207008)
```

Scale covariates.
```{r}
head(data_hmm)
data_hmm$rugged49_scale <- scale(data_hmm$rugged49.clean)
data_hmm$rugged25_scale <- scale(data_hmm$rugged25.clean)
data_hmm$rugged9_scale <- scale(data_hmm$rugged9.clean)
data_hmm$hq_scale <- scale(data_hmm$hq_dist)
data_hmm$view_scale <- scale(data_hmm$view_for_kg_proj)
data_hmm$road_scale <- scale(data_hmm$road.dist.clean)
data_hmm$grass_scale <- scale(data_hmm$grass_120m)
data_hmm$chap_scale <- scale(data_hmm$chap_120m)
data_hmm$wood_scale <- scale(data_hmm$wood_120m)
head(data_hmm)
```


Look at step lengths and turn angles.
```{r}
hist(data_hmm$step)
hist(data_hmm$angle)

# what about the small steps? zoom in
data_hmm_small <- data_hmm %>% 
  filter(step < 0.15)
hist(data_hmm_small$step)

# determine proportion of step lengths equal to 0
whichzero <- which(data_hmm$step == 0)
length(whichzero)/nrow(data_hmm)
```


## TWO STATE MODELS

Determine and set initial parameters for mu0 and sigma0 based on observation of histograms.
See this guide for setting parameters: https://cran.r-project.org/web/packages/moveHMM/vignettes/moveHMM-starting-values.pdf
```{r}
mu0_2state <- c(0.01, 0.3) # step mean (two parameters: one for each state)
sigma0_2state <- c(0.01, 0.3) # step SD
zeromass0_2state <- c(0.50, 0.10) # step zero-mass
stepPar0_2state <- c(mu0_2state,sigma0_2state,zeromass0_2state)

angleMean0_2state <- c(0,pi) # angle mean
kappa0_2state <- c(1,1) # angle concentration
anglePar0_2state <- c(angleMean0_2state,kappa0_2state)

angleMean0_2state <- c(0,0) # angle mean
kappa0_2state <- c(0.5,1) # angle concentration
anglePar0_2state <- c(angleMean0_2state,kappa0_2state)
```

### Null model
```{r}
model0 <- fitHMM(data=data_hmm, 
                 nbStates=2, 
                 stepPar0=stepPar0_2state, 
                 anglePar0=anglePar0_2state)

model0 # model summary
plot(model0) # hit escape to get out of the long list

# proportion of time spent in each state
states <- viterbi(model0)
prop.table(table(states)) 

# plot states for a given individual
plotStates(model0, animals="080815_01")

AIC(model0) # AIC = 105,028
```


### Ruggedness
```{r}
model1 <- fitHMM(data=data_hmm, 
                 nbStates=2, 
                 stepPar0=stepPar0_2state, 
                 anglePar0=anglePar0_2state,
                 formula = ~rugged25_scale)
model1 # model summary
AIC(model1) # AIC = 104937.6

model2 <- fitHMM(data=data_hmm, 
                 nbStates=2, 
                 stepPar0=stepPar0_2state, 
                 anglePar0=anglePar0_2state,
                 formula = ~rugged49_scale)
model2 # model summary
AIC(model2) # AIC = 104971.9

model3 <- fitHMM(data=data_hmm, 
                 nbStates=2, 
                 stepPar0=stepPar0_2state, 
                 anglePar0=anglePar0_2state,
                 formula = ~rugged9_scale)
model3 # model summary
AIC(model3) # AIC = 104893.4 - best with rugged9
```

### Viewshed
```{r}
model4 <- fitHMM(data=data_hmm, 
                 nbStates=2, 
                 stepPar0=stepPar0_2state, 
                 anglePar0=anglePar0_2state,
                 formula = ~view)
model4 # model summary
AIC(model4) # AIC = 104864.9
```

### Distance to road
```{r}
model5 <- fitHMM(data=data_hmm, 
                 nbStates=2, 
                 stepPar0=stepPar0_2state, 
                 anglePar0=anglePar0_2state,
                 formula = ~road.dist.clean)
model5 # model summary
AIC(model5) # AIC = 103029.9
```

### Vegetation type
Probably shouldn't use—this is a numeric variable, but really should be treated as categorical (and model doesn't allow for categorical variables)
```{r}
model6 <- fitHMM(data=data_hmm, 
                 nbStates=2, 
                 stepPar0=stepPar0_2state, 
                 anglePar0=anglePar0_2state,
                 formula = ~vegetation.coarser.clean2)
model6 # model summary
AIC(model6) # AIC = 104875.3
```

### Chaparral within 120 meters
```{r}
model7 <- fitHMM(data=data_hmm, 
                 nbStates=2, 
                 stepPar0=stepPar0_2state, 
                 anglePar0=anglePar0_2state,
                 formula = ~chap_120m)
model7 # model summary
AIC(model7) # AIC = 105027
```

### Grassland within 120 meters
```{r}
model8 <- fitHMM(data=data_hmm, 
                 nbStates=2, 
                 stepPar0=stepPar0_2state, 
                 anglePar0=anglePar0_2state,
                 formula = ~grass_120m)
model8 # model summary
AIC(model8) # AIC = 104916.7
```

### Woodland within 120 meters
```{r}
model9 <- fitHMM(data=data_hmm, 
                 nbStates=2, 
                 stepPar0=stepPar0_2state, 
                 anglePar0=anglePar0_2state,
                 formula = ~wood_120m)
model9 # model summary
AIC(model9) # AIC = 104898.5
```

### Distance to headquarters
```{r}
model10 <- fitHMM(data=data_hmm, 
                 nbStates=2, 
                 stepPar0=stepPar0_2state, 
                 anglePar0=anglePar0_2state,
                 formula = ~hq_dist)
model10 # model summary
AIC(model10) # AIC = 104948
```

Compare the two-state models
```{r}
AIC(model0, model1, model2, model3, model4, model5, model6, model7, model8, model9, model10)
# best is model 5 = road.dist.clean
```